---
title: "Swimmer_Plots"
author: "Mackenzie Mekler"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE, message = FALSE)
```


## Load Libraries : Need to be run for every new session
```{r, message = FALSE}
library(tidyverse) #ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
library(swimplot)

# Set global ggplot theme
theme_set(theme_grey() + theme(text = element_text(size = 24), legend.position="bottom"))
```


## Load and Modify the csv file
```{r, message = FALSE}
kms <- read_csv("test.csv")
kms$`Field Point` <- factor(kms$`Field Point`)

# for testing 
# kms <- kms[7,]
```


## Number of T cells attached
```{r, message = FALSE}
vecs <- list()
for(cell in 1:nrow(kms)){
  tcellVec <- replicate(97, 0)
  # print(kms[cell,])
  cnx1 <- kms[cell,]$`#1o.Connection`
  cnx2 <- kms[cell,]$`#2o.Connection`
  cnx3 <- kms[cell,]$`#3o.Connection`
  
  # I can make a vector of length 97 and put either 0, 1, 2, or 3 based on number of T cells 
  # another vector can be 0, 1, 2, 3 based on nothing, lost, av+, or tcav+
  
  # for cnx1
  if(grepl(";", cnx1)){
    split1a <- str_split(cnx1, ';')
    for(item in split1a[[1]]){
      # print(item)
      mysplit <- strsplit(item, '-')
      start <- as.integer(mysplit[[1]][1])
      end <- as.integer(mysplit[[1]][2])
      if(length(mysplit[[1]]) < 2){
        end <- start
      }
      for(i in 1:length(tcellVec)){
        if(start <= i & i <= end){
          tcellVec[i] <- tcellVec[i] + 1
        }
      }
    }
  }
  else {
    mysplit <- strsplit(cnx1, '-')
    start <- as.integer(mysplit[[1]][1])
    end <- as.integer(mysplit[[1]][2])
    if(length(mysplit[[1]]) < 2){
      end <- start
    }
    for(i in 1:length(tcellVec)){
      if(start <= i & i <= end){
        tcellVec[i] <- tcellVec[i] + 1
      }
    }
  }
  
  # for cnx2
  if(!is.na(cnx2))
  if(grepl(";", cnx2)){
    split1a <- str_split(cnx2, ';')
    for(item in split1a[[1]]){
      # print(item)
      mysplit <- strsplit(item, '-')
      start <- as.integer(mysplit[[1]][1])
      end <- as.integer(mysplit[[1]][2])
      if(length(mysplit[[1]]) < 2){
        end <- start
      }
      for(i in 1:length(tcellVec)){
        if(start <= i & i <= end){
          tcellVec[i] <- tcellVec[i] + 1
        }
      }
    }
  }
  else {
    mysplit <- strsplit(cnx2, '-')
    start <- as.integer(mysplit[[1]][1])
    end <- as.integer(mysplit[[1]][2])
    if(length(mysplit[[1]]) < 2){
      end <- start
    }
    for(i in 1:length(tcellVec)){
      if(start <= i & i <= end){
        tcellVec[i] <- tcellVec[i] + 1
      }
    }
  }
  
  # for cnx3
  if(!is.na(cnx3))
  if(grepl(";", cnx3)){
    split1a <- str_split(cnx3, ';')
    for(item in split1a[[1]]){
      # print(item)
      mysplit <- strsplit(item, '-')
      start <- as.integer(mysplit[[1]][1])
      end <- as.integer(mysplit[[1]][2])
      if(length(mysplit[[1]]) < 2){
        end <- start
      }
      for(i in 1:length(tcellVec)){
        if(start <= i & i <= end){
          tcellVec[i] <- tcellVec[i] + 1
        }
      }
    }
  }
  else {
    mysplit <- strsplit(cnx3, '-')
    start <- as.integer(mysplit[[1]][1])
    end <- as.integer(mysplit[[1]][2])
    if(length(mysplit[[1]]) < 2){
      end <- start
    }
    for(i in 1:length(tcellVec)){
      if(start <= i & i <= end){
        tcellVec[i] <- tcellVec[i] + 1
      }
    }
  }
  vecs[[cell]] <- tcellVec
}


```









## Final Vectors 
```{r}
for(i in 1:length(vecs)){
  r <- kms[i,]
  if(!is.na(r$`Myelo AnnexinV+`)){
    av <- r$`Myelo AnnexinV+`
    vecs[[i]][as.integer(av)] <- 4
  }
  if(!is.na(r$`Tcell AnnexinV+`)){
    tcav <- str_split(r$`Tcell AnnexinV+`, ',')
    for(num in 1:length(tcav)){
      vecs[[i]][as.integer(tcav[[1]][num])] <- 5
    }
  }
  if(!is.na(r$`Stack # (lost)`)){
    lost <- r$`Stack # (lost)`
    vecs[[i]][as.integer(lost)] <- 6
  }
}

```



https://kathoffman.github.io/swimmer-plots/slides-wcm.html#6
## Make the graphs 

note that the fig.width and fig.height in the heading are how I have been adjusting the size of the graph
I found that it was easier to do it this way than in the ggplot itself
```{r fig.align="center", echo=FALSE, fig.width=15, fig.height=30}

# I think I need to change the way that my data is stored in the final df to be more like in the hyperlink

final_vecs <- data.frame(
  status = unlist(vecs), 
  id = rep(seq_along(vecs), sapply(vecs, length))
)

final_vecs$timepoint <- rep(1:97, length.out = nrow(final_vecs))
final_vecs$status <- factor(final_vecs$status, levels = c("0", "1", "2", "3", "4", "5", "6"))
final_vecs$id <- as.factor(final_vecs$id)

# final_vecs <- as.data.frame(do.call(rbind, vecs))
# final_vecs$id <- 1:length(kms$`Cell#`)

final_vecs %>%
  # filter(as.integer(id) < 10) %>%
  ggplot(aes(y=id, x=timepoint, color=status)) +
  theme_bw() +
  geom_line(size=3, aes(group=id), na.rm = TRUE) + 
  scale_color_manual(values = c(
    "0" = "lightgreen", 
    "1" = "lightblue",
    "2" = "blue",
    "3" = "darkblue", 
    "4" = "red",
    "5" = "maroon",
    "6" = "darkgray"
  ))


  # +
  # geom_point(aes(x=av)) +
  # geom_point(aes(x=severe_this_day)) +
  # geom_point(aes(x=death_this_day))
```

In the future I will be able to add segements of code in here that pick out the specific values that I want to show much more easily than using excel. 